---
title: "Lab 08 - Text Mining/NLP"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(eval = F, include  = T)
```
<a href="https://github.com/CuriousChum/JSC370-labs/blob/main/lab8/08lab-text-mining.md" target="_blank">link to GitHub repo</a>

# Learning goals

- Use `unnest_tokens()` and `unnest_ngrams()` to extract tokens and ngrams from text
- Use dplyr and ggplot2 to analyze and visualize text data
- Try a theme model using `topicmodels`

# Lab description

For this lab we will be working with the medical record transcriptions from https://www.mtsamples.com/ available at https://github.com/JSC370/JSC370-2025/tree/main/data/medical_transcriptions.

# Deliverables

1. Questions 1-7 answered, knit to pdf or html output uploaded to Quercus.

2. Render the Rmarkdown document using `github_document` and add it to your github site. Add link to github site in your html.


### Setup packages

You should load in `tidyverse`, (or `data.table`), `tidytext`, `wordcloud2`, `tm`, and `topicmodels`.


## Read in the Medical Transcriptions

Loading in reference transcription samples from https://www.mtsamples.com/

```{r eval=FALSE}
library(tidytext)
library(tidyverse)
library(wordcloud2)
library(tm)
library(topicmodels)

mt_samples <- read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/main/data/medical_transcriptions/mtsamples.csv")
mt_samples <- mt_samples |>
  select(description, medical_specialty, transcription)

head(mt_samples)
```

---

## Question 1: What specialties do we have?

We can use `count()` from `dplyr` to figure out how many different medical specialties are in the data. Are these categories related? overlapping? evenly distributed? Make a bar plot.

```{r eval=FALSE}
mt_samples |>
  count(medical_specialty, sort = TRUE) |>
  head(20) |>
  ggplot(aes(fct_reorder(medical_specialty, n), n)) +
  coord_flip() +
  geom_col(fill='blue') 
```

---

## Question 2: Tokenize

- Tokenize the the words in the `transcription` column
- Count the number of times each token appears
- Visualize the top 20 most frequent words with a bar plot
- Create a word cloud of the top 20 most frequent words

### Explain what we see from this result. Does it makes sense? What insights (if any) do we get?

```{r eval=FALSE}
tokens <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription) |>
  group_by(word) |>
  summarize(word_frequency=n()) |>
  arrange(across(word_frequency, desc)) |>
  head(20)

tokens |>
  ggplot(aes(fct_reorder(word, word_frequency),word_frequency)) +
  geom_bar(stat='identity', fill='dodgerblue') + # eqv to geom_col (?)
  coord_flip() +
  theme_bw()
tokens |>
  count(word, sort=TRUE) |>
  wordcloud2(size=0.4, color='random-light')
```

---

## Question 3: Stopwords

- Redo Question 2 but remove stopwords
- Check `stopwords()` library and `stop_words` in `tidytext`
- Use regex to remove numbers as well
- Try customizing your stopwords list to include 3-4 additional words that do not appear informative

### What do we see when you remove stopwords and then when you filter further? Does it give us a better idea of what the text is about?

```{r eval=FALSE}
head(stopwords("english"))
length(stopwords("english"))
head(stop_words)

tokens <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription, token='words') |>
  group_by(word) |>
  summarize(word_freq=n()) |>
  arrange(across(word_freq, desc)) |>
  anti_join(stop_words, by='word') |>
  filter(!str_detect(word, "^\\d+$")) |>
  filter(!word %in% c("mm", "mg", "noted")) |> # additional stopwords
  top_n(20, word_freq)
  
tokens |>
  ggplot(aes(fct_reorder(word, word_freq), word_freq)) +
  xlab("tokens") +
  coord_flip() +
  geom_bar(stat='identity', fill='dodgerblue')

tokens |>
  wordcloud2(size=0.7, color='random-light')
```


---



## Question 4: ngrams

Repeat question 2, but this time tokenize into bi-grams. How does the result change if you look at tri-grams? Note we need to remove stopwords a little differently. You don't need to recreate the wordclouds.

```{r eval=FALSE}
sw2 <- c(stop_words$word, "mm", "mg", "noted")
sw_pattern <- paste0("(^| )(", paste(sw2, collapse="|"), ")( |$)")

tokens_bigram <- mt_samples |>
  select(transcription) |>
  unnest_tokens(ngram, transcription, token = "ngrams", n = 2) |>
  filter(!grepl(sw_pattern, ngram, ignore.case=TRUE))|>
  filter(!grepl("(^| )[0-9.]+( |$)", ngram))|>
  group_by(ngram) |>
  summarize(word_freq=n()) |>
  arrange(across(word_freq, desc))

write_csv(tokens_bigram, 'bigrams.csv')

tokens_bigram |>
  top_n(20, word_freq) |>
  ggplot(aes(word_freq, ngram)) +
  geom_col(fill='dodgerblue') +
  theme_bw()
  ... 
```
---

## Question 5: Examining words

Using the results from the bigram, pick a word and count the words that appear before and after it, and create a plot of the top 20.

```{r eval=FALSE}
library(stringr)
# e.g. patient, blood, preoperative...
chosen <- "blood"
tokens_bigram |>
  filter(str_detect(ngram, '(^| )blood( |$)')) |>
  mutate(word = str_remove(ngram, "blood"),
    word = str_remove_all(word, " ")) |>
  group_by(word) |>
  head(20) |>
  ggplot(aes(word_freq, fct_reorder(word, word_freq))) +
  geom_col(fill='dodgerblue') +
  theme_bw()
  
      
```

---


## Question 6: Words by Specialties

Which words are most used in each of the specialties? You can use `group_by()` and `top_n()` from `dplyr` to have the calculations be done within each specialty. Remember to remove stopwords. How about the 5 most used words?


```{r eval=FALSE}
topw_by_spec <- mt_samples |>
  unnest_tokens(word, transcription, token='words') |>
  anti_join(stop_words, by='word') |>
  filter(!str_detect(word, "^\\d+$")) |>
  filter(!word %in% c("mm", "mg", "noted")) |>
  count(medical_specialty, word, sort = TRUE) |>
  group_by(medical_specialty) |>
  top_n(3) |>
  ungroup()

topw_by_spec |>
  ggplot(aes(reorder_within(word, n, medical_specialty), n,
             fill=medical_specialty)) +
  geom_col(show.legend=FALSE) +
  facet_wrap(~medical_specialty, scales='free_y', ncol=5) +
  scale_x_reordered() +
  coord_flip() +
  theme_bw()
```


## Question 7: Topic Models

See if there are any themes in the data by using a topic model (LDA). 

- you first need to create a document term matrix
- then you can try the LDA function in `topicmodels`. Try different k values.
- create a facet plot of the results from the LDA (see code from lecture)


```{r eval=FALSE}

transcripts_dtm <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription) |>
  anti_join(stop_words, by="word") |>
  filter(!str_detect(word, "^\\d+$")) |>
  filter(!word %in% c("mg", "mm", "noted")) |>
  DocumentTermMatrix()

transcripts_dtm <- as.matrix(transcripts_dtm)   

transcripts_lda <- LDA(transcripts_dtm, k=3, control=list(seed=777))

transcript_top_terms <- tidy(transcripts_lda, matrix='beta') |>
  group_by(topic) |>
  top_n(5, beta) |>
  ungroup() |>
  arrange(topic, desc(beta))

transcript_top_terms

transcript_top_terms |>
  ggplot(aes(reorder_within(term, beta, topic), beta,
             fill = as.factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~topic) +
  scale_x_reordered() +
  coord_flip() +
  theme_bw()
```
```{r}
rmarkdown::render("08lab-text-mining.Rmd", output_format = "github_document")
```




